Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-12-23 paper/manuscript.tex
Tweaks from @mrshirts

Alchemical free energy calculations are a useful tool for predicting free energy differences associated with the transfer of molecules from one environment to another.
The hallmark of these methods is the use of ``bridging'' potential energy functions representing \emph{alchemical} intermediate states that cannot exist as real chemical species. The data collected from these bridging alchemical thermodynamic states allows the efficient computation of transfer free energies (or differences in transfer free energies) with orders of magnitude less simulation time than simulating the transfer process directly. 
While these methods are highly flexible, care must be taken in avoiding common pitfalls to ensure that computed free energy differences can be robust and reproducible for the chosen force field, and that appropriate corrections are included to permit direct comparison with experimental data.

In this paper, we review current best practices for several popular application domains of  alchemical free energy calculations performed with equilibrium simulations, in particular relative and absolute small molecule binding free energy calculations to biomolecular targets.

@@

\section{What are alchemical free energy methods?}
\label{sec:intro}
Alchemical free energy calculations compute free energy differences associated with transfer processes, such as the binding of a small molecule to a receptor, the transfer of a small molecule from an aqueous to apolar phase~\cite{zwanzig1954hightemperature}, or the effects of protein side chain mutations on binding affinities or thermostabilities. 
 These calculations use non-physical\footnote{Here, the non-physical nature of the transformation is referred to as "alchemical", a term coined by Tembre and McCammon in Ref.~\cite{tembre1984ligandreceptor}.} intermediate states in which the chemical identity of some portion of the system (such as a small molecule ligand or protein sidechain) is changed by modifying the potential governing the interactions with the environment for the atoms being modified, inserted, or deleted.

Fig.~\ref{fig:fig_what_is_alchemy} illustrates common free energy changes that may be difficult to compute with unbiased molecular dynamics methods, but are more tractable with alchemical methods.
In alchemical simulations, the introduction of intermediate \textit{alchemical states} that bridge the high-probability regions of configuration space between two physical endstates of interest, permits the robust computation of free energy for large transformations.
Alchemical calculations can be used in a variety of scenarios, such as: 
\begin{itemize}
\item computing the free energy of a conformational change for a molecule with a high barrier to interconversion (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{B});
\item computing partition ($\log P$) or distribution ($\log D$) coefficients between environments (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{A})~\cite{rustenburg2016measuring, bosisio2016blinded} 
\item determining partitioning between compartments into membranes (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{D})~\cite{corey2019insights}. 
\end{itemize}

Furthermore, alchemical calculations are frequently used to estimate changes in free energies upon modifying a ligand or protein: 
\begin{itemize}
\item a protein residue can be alchemically mutated to probe the impact on binding affinity (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{F})\cite{hauser2018predicting,aldeghi2018accurate} or changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate}; 
\item the entire ligand can be alchemically transferred from protein to solvent in an absolute binding free energy calculation (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{C})~\cite{mobley2007predicting,aldeghi2015accurate,aldeghi2017predictions}; 
\item small alchemical modifications can be made between chemically related ligands to estimate relative differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{E})~\cite{wang2015accurate,mey2016blinded,song2019using,gapsys2020large,kuhn2020assessment}.
\end{itemize}

After an alchemical calculation is performed, which generally involves multiple simulations at a variety of alchemical states, the data must be analyzed to compute an estimate of the free energy for the transformation of interest.
 Early work used simple but statistically suboptimal estimators for this: free energy perturbation (FEP) used a simple (but highly biased) estimator based on the Zwanzig relation~\cite{zwanzig1954hightemperature} or numerical quadrature via thermodynamic integration (TI), for which the theory dates back the better part of a century but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamics, merz1989free}. Macuglia et al. review the evolution of biomolecular alchemical free energy calculations into practically useful tools \cite{macuglia2025dawn}. %
 More recent developments have seen new, highly efficient statistical estimators that make better use of all the data, often building on the more efficient and less biased Bennett acceptance ratio (BAR)~\cite{bennett1976efficient}, producing multistate generalizations~\cite{shirts2008statisticallya} or removing the need for global equilibrium~\cite{wu2016multiensemble, mey2014xtram, wu2014statistically}.

@@

As the field of molecular simulation can now routinely access microsecond timescales with the aid of GPUs~\cite{salomon-ferrer2013routine}, and millisecond timescales appear to soon be within reach, accurate alchemical calculations on even more challenging problems will become reasonable to perform. 
In the meantime, today's users may find it difficult to get started with these complex calculations whilst also keeping up with the fast pace of change. 
This Best Practices guide provides current recommendations and tips for users of all experience. Updates and suggestions are welcomed via our GitHub repository at \url{https://github.com/alchemistry/alchemical-best-practices}.

@@

This Best Practices guide focuses on providing a good starting point for new practitioners and a reference for experienced practitioners. 
 For this purpose we provide a convenient checklist (Sec.~\ref{sec:checklist}) to help ensure all calculations can take advantage of currently-understood best practices for alchemical simulation and analysis. Where the best practices are currently not certain, we highlight areas where further research is needed to identify an unambiguous recommendation.
 This guide can also serve as a set of best practices to ensure simulation robustness and reproducibility which reviewers may wish to consider as they evaluate papers including alchemical free energy calculations.

In this best practices document, we assume that novice practitioners have at least moderate experience with molecular simulation concepts and use of simulation packages. 
We also assume basic familiarity with the principles of molecular mechanics, molecular dynamics simulations, statistical mechanics, and the biophysics of protein-ligand association. If you feel unfamiliar with some of these concepts, good starting points can be found in these references~\cite{braun2019best, grossfield2018best, klimovich2015guidelines, shirts2012best}.

 While reading this Best Practices guide, it is important to bear in mind \emph{this is not a review} of all free energy calculation methods at the cutting edge of current research.
Instead this guide aims to answer the following questions:
\begin{itemize}
    \item Is my problem suitable for an alchemical calculation? 
     \item How do I select an appropriate simulation protocol? 
     \item What software tools are available to perform alchemical calculations? 
     \item How should I analyze my data and report uncertainties? 
\end{itemize}

Notably, this document---version 2.0 of the original guide---focuses specifically on best practices on single alchemical free energy calculations. A companion guide, Best Practices for Multiple Alchemical Free Energy Calculations, covers the considerations involved in high-throughput or large-scale campaigns, such as automation, perturbation network design, and campaign-level data management and analysis. Readers aiming to scale up to multiple calculations may find that guide more relevant to those aspects.

In this guide, some other background information may be needed depending on the nature of the alchemical project. For example, often, if binding poses are not known, docking calculations or co-folding models can be used to generate an initial small molecule binding pose to start alchemical simulations\cite{scheen2025leveraging, thaler2025boltz}. This will require some basic familiarity on how to perform docking or using generative models. to generate reasonable simulation starting points~\cite{grinter2014challenges}.

As some of the theoretical background can seem daunting, we do, however, provide a guide to the essential theory behind alchemical free energy calculations in Sec.~\ref{sec:theory}.
In the remainder of this paper, we will cover topics that are key to the preparation (Sec.~\ref{sec:prerequisites}), choice and use of correct protocols (Sec.~\ref{sec:simulation_protocol_choice}), and finally the best practices that should be used in the analysis of alchemical calculations (Sec.~\ref{sec:data_analysis}). 
We will give particular focus to aspects of the molecular simulations which are unique to alchemical calculations---these include the calculation of transfer free energies (hydration free energies, partition coefficients, etc.), and binding free energies (absolute and relative). We primarily focus on free energy calculations using simulations performed at \emph{equilibrium} in this Best Practices guide, as best practices for these are more developed, and non-equilibrium techniques may warrant their own guide as such practices evolve.

While we try to address as many methods and practices as possible, the field of free energy calculations is broad, and there are many advanced topics that are left to future Best Practices documents focusing on specific issues. 
Below, we provide a non-exhaustive list of topics we have \emph{not} addressed, along with some references to provide starting points on these more advanced topics:
\begin{itemize}
\item covalent inhibition~\cite{lameira2019predicting}
\item free energies of mutation of protein side chains~\cite{gapsys2016accurate,aldeghi2018accurate}
\item nonspecific binding or multiple binding sites~\cite{gill2018binding}
\item approximate and often less accurate endpoint free energy methods such as MM-PBSA~\cite{genheden2015mm} and LIE~\cite{gutierrez-de-teran2012linear}
\item Free energy methods that extract the ligand using geometric order parameters and potential of mean force methods~\cite{heinzelmann2017attachpullrelease}
\item forcefield dependence for protein, ligand, ions, co-solvents, and co-factors. A number of different studies have looked at the influence of force fields and it is assumed the user has made an appropriate choice for the system under study~\cite{loeffler2018reproducibility, vassetti2019assessment, lopes2015current}. 
\item non-equilibrium free energy calculations~\cite{gapsys2020large}
\item Free energy calculations using QM/MM methods~\cite{beierlein2011simple,dybeck2016comparison,cave-ayland2015direct,Schoeller2023}.
\item Free energy calculations using machine learning methods~\cite{rufa2020chemical, scheen2020hybrid, cole2020machine,Karwounopoulos2024}
\end{itemize}

@@

\begin{tcolorbox}[title=List of Symbols, colback=green!10!white]
$A$ --- Helmholtz free energy (free energy in the canonical ensemble) or Helmholtz function, with Helmholtz free energy used in the text.\\
$b$ --- reduced binding energy function \\
$C^{\circ}$ --- standard state concentration \\
$C_t$ --- discrete-time-normalized fluctuation auto-correlation function\\
$f$ --- reduced (dimensionless) free energy \\
$G$ --- Gibbs free energy (free energy in the isothermal isobaric ensemble), Gibbs function, or free enthalpy, though the most common term Gibbs free energy is used in the text\\
$g$ --- statistical inefficiency\\
$h$ --- Planck's constant \\
$\vec{h}$ --- a displacement vector in the ATM method
$k_B$ --- Boltzmann constant \\
$k_H$ --- Henry's constant \\
$K_b^{\circ}$ --- binding  constant \\
$L$ and $R$ --- generic names for ligand and receptor\\
$\mathcal{O}$ --- overlap matrix\\
$P^{\circ}$ --- standard pressure \\
$p$ --- pressure \\
$\vec{q}$ --- vector of a single configuration, i.e. $x$, $y$, $z$ coordinates of the simulation system\\
$T$ --- temperature \\
$t_0$ --- equilibration time \\
$U$ --- potential energy\\
$u$ --- reduced (dimensionless) potential describing a thermodynamic state \\
$w(b, \vec{\lambda})$ --- perturbation energy function \\
$Z$ --- partition function \\
$\beta \equiv (k_B T)^{-1}$ --- inverse thermal energy \\
$\Gamma$ --- configurational space accessible by simulations \\
$\Delta \hat{f}$ --- estimate from an estimator for the reduced free energy difference between two states\\
$\vec{\lambda}$ --- alchemical progress parameter, which may be multidimensional \\
$\mu$ --- chemical potential (grand canonical ensemble)\\
$\tau _{eq}$ --- integrated auto-correlation time\\
\end{tcolorbox}

@@

In this section, we use a statistical mechanics theory of dilute solutions to derive free energy expressions that form the basis of alchemical computational protocols to model non-covalent molecular binding, solvation, and solvent partitioning equilibria. The emphasis here is placed on bridging theoretical foundations and intuition. Since chemical equilibria are regulated by the chemical potentials of the species involved, we will start by considering a statistical mechanics expression of the chemical potential of a solute in a solvent. We will then combine chemical potentials to develop models for each process.

@@

The chemical potential of a solute $u$ in an ideally diluted solution in a solvent $v$ can be written as~\cite{guggenheim1952mixtures504,gilson1997statisticalthermodynamic,gilson2007calculation}
\begin{equation}
  \mu_u = -k_B T \ln \frac{q'_u}{C_u \Lambda_u^3}
  \label{eq:chemical-potential-definition}
\end{equation}
where $k_B$ is Boltzmann's constant, $T$ is the absolute temperature, $C_u = N_u/V$ is the number concentration of the solute, $\Lambda_u = h/\sqrt{2 \pi M_u k_B T}$ is the thermal De Broglie wavelength of the solute, where $h$ is Planck's constant and $M_u$ is the solute's total mass. The quantity $q'_u$ in Eq.~(\ref{eq:chemical-potential-definition}) is the molecular partition function of the solute at rest in the solvent
\begin{equation}
  q'_u = \frac{\Lambda_u^3}{\prod_i \lambda_i^3} 8 \pi^2 z_u
  \label{eq:intra-qpu-def}
\end{equation}
where $\beta = 1/(k_B T)$, $\lambda_i$ is the thermal De Broglie wavelength of the $i$-th atom of the solute, and $z_u$ is the intramolecular configurational partition function of the solute
\begin{equation}
  z_u = \int dx_u e^{-\beta \Psi_v(x_u)}
  \label{eq:intra-zu-def}
\end{equation}
where $x_u$ represents the collection of internal degrees of freedom of the solute (bond lengths, bond angles, and dihedral angles) with the corresponding volume element $dx_u$ assumed to include the appropriate Jacobian factors. The function $\Psi_v(x_u)$ in Eq.~\ref{eq:intra-zu-def} is the solvent-averaged potential of mean force of the solute in conformation $x_u$ in solvent $v$\cite{rouxandsimonson1999solventmodels} 
\begin{equation}
  e^{-\beta \Psi_v(x_u)} = \frac{1}{Z_{N_v}} \int d\vec{r}_v e^{-\beta U(x_u, \vec{r}_v)}
  \label{eq:solvent-pmf-def}
\end{equation}
where $\vec{r}_v$ represents the collection of Cartesian coordinates of the atoms of $N_v$ solvent molecules, $U(x_u, \vec{r}_v)$ is the potential energy of the solvent with one solute molecule in conformation $x_u$ placed in an arbitrary position in a volume $V$. When the potential energy function of the solution is separable into an intramolecular potential energy of the solute $U(x_u)$ and a solute-solvent non-bonded interaction energy, the potential of mean force can be written as the sum $\Psi_v(x_u) = U(x_u) + W_v(x_u)$ where $W_u(x_u)$ is the solvent potential of mean force, the solvation free energy of the solute fixed in configuration $x_u$.\cite{gilson2007calculation} However, the potential of mean force definition Eq.~(\ref{eq:solvent-pmf-def}) is valid in general. Finally, $Z_{N_v}$ is the configurational partition function of the pure solvent
\begin{equation}
  Z_{N_v} = \int d\vec{r}_v e^{-\beta U(\vec{r}_v)} \, .
  \label{eq:Z-pure-solvent}
\end{equation}

A few notes are in order. First, Eq.~(\ref{eq:chemical-potential-definition}) is derived using classical statistical mechanics theory.~\cite{hill1986statthermo} Classical mechanics is considered an excellent approximation for the room temperature processes studied here involving only changes in intermolecular interactions. A quantum mechanical treatment is necessary for processes not covered here, such as covalent binding, that involve breaking and forming chemical bonds. Classically, equilibrium thermodynamics properties, such as gas solubilities and binding constants, do not depend on atomic masses. Hence, while they are included in Eqs.~\ref{eq:chemical-potential-definition} and Eq.~\ref{eq:intra-qpu-def} for completeness, the thermal De Broglie wavelength terms coming from the partition functions of the momenta cancel in the resulting expressions below and we will no longer consider them.

Eq.~(\ref{eq:chemical-potential-definition}) is derived by writing the classical canonical configurational partition function $Z(N_u = 1, N_v)$ of one solute molecule in a solvent of $N_v$ molecules in a volume $V$ and multiplying and dividing it by configurational partition function of the pure solvent [Eq.~(\ref{eq:Z-pure-solvent})], and separating out the integration over the external degrees of freedom of the solute (translations and rotations) giving an $8 \pi^2 V$ term to yield $Z(N_u = 1, N_v) = Z_{N_v} 8 \pi^2 V z_u$ where $z_u$ is defined by Eqs.~(\ref{eq:intra-zu-def}) and (\ref{eq:solvent-pmf-def}). The canonical configurational partition function of $N_u$ independent solute molecules in a solvent is then obtained by considering the product of the solute terms $Z(N_u, N_v) = Z_{N_v} (8 \pi^2 V z_u)^{N_u}/N_u!$,~\cite{guggenheim1952mixtures504,simonson2016physical} where the factorial term takes care of indistinguishability. Finally, the canonical partition function $Q(N_u, N_v)$ of the solution is obtained by including the appropriate terms coming from the momenta and differentiating the resulting Helmholtz free energy $A(N_u, N_v) = -k_B T \ln Q(N_u, Nv)$ with respect to $N_u$ to obtain Eq.~(\ref{eq:chemical-potential-definition}).

@@

We derived Eq.~(\ref{eq:chemical-potential-definition}) using the canonical (constant volume) partition function of the solution. Since the chemical potential can be equivalently obtained from the constant-volume differentiation of the Helmholtz free energy or the constant-pressure differentiation of the Gibbs free energy, Eq.~(\ref{eq:chemical-potential-definition}) is correct. However, the volume dependence of the chemical potential has to be considered when modeling the Gibbs free energy change for constant-pressure processes where the system's volume changes significantly.~\cite{gilson1997statisticalthermodynamic} Terms due to volume variations are generally considered small for most processes of solvation, solvent partitioning, and molecular binding at atmospheric pressure and are often ignored in alchemical calculations. Hence, we will use Eq.~(\ref{eq:chemical-potential-definition}) to derive expressions for $\Delta G$'s.

>>>>

Alchemical free energy calculations are a useful tool for predicting free energy differences associated with the transfer of molecules from one environment to another.
The hallmark of these methods is the use of ``bridging'' potential energy functions representing \emph{alchemical} intermediate states that cannot exist as real chemical species. The data collected from these bridging alchemical thermodynamic states allows the efficient computation of transfer free energies (or differences in transfer free energies) with orders of magnitude less simulation time than simulating the transfer process directly. 
While these methods are highly flexible, significant care must be taken in avoiding common pitfalls to ensure that computed free energy differences can be robust and reproducible for the chosen force field, and that appropriate corrections are included to permit direct comparison with experimental data.

In this paper, we review current best practices for several popular application domains of  alchemical free energy calculations performed with molecular dynamics simulations, with a particular focus on the intricacies and challenges of relative and absolute free energy calculations of binding of small molecules to biomolecular targets.

@@

\section{What are alchemical free energy methods?}
\label{sec:intro}
Alchemical free energy calculations allow the estimation of free energy differences associated with transfer processes, such as the binding of a small molecule to a receptor, the transfer of a small molecule from an aqueous to apolar phase~\cite{zwanzig1954hightemperature}, or the effects of protein side chain mutations on binding affinities or thermostabilities. 
 These calculations use non-physical\footnote{Here, the non-physical nature of the transformation is referred to as "alchemical", a term coined by Tembre and McCammon in Ref.~\cite{tembre1984ligandreceptor}.} intermediate states in which the chemical identity of some portion of the system (such as a small molecule ligand or protein side chain) is changed by modifying the potential governing the interactions with the environment for the atoms being modified, inserted, or deleted.

Fig.~\ref{fig:fig_what_is_alchemy} illustrates common free energy changes that may be difficult to compute with unbiased molecular dynamics methods, but are more tractable with alchemical methods.
In alchemical simulations, the introduction of intermediate \textit{alchemical states} that bridge the high-probability regions of configuration space between two physical endstates of interest permits the robust computation of free energy for large transformations.
Alchemical calculations can be used in a variety of scenarios, such as: 
\begin{itemize}
\item computing the free energy of a conformational change for a molecule with a high barrier to interconversion (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{B});
\item computing partition ($\log P$) or distribution ($\log D$) coefficients of a molecule moving between different environments (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{A})~\cite{rustenburg2016measuring, bosisio2016blinded} 
\item determining the free energy of partitioning between compartments into membranes (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{D})~\cite{corey2019insights}. 
\end{itemize}

Furthermore, alchemical calculations are frequently used to estimate changes in free energies upon modifying a ligand or protein. Specifically:
\begin{itemize}
\item a protein residue can be alchemically mutated to probe the impact on binding affinity (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{F})\cite{hauser2018predicting,aldeghi2018accurate} or changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate}; 
\item the entire ligand can be alchemically transferred from protein to solvent in an absolute binding free energy calculation (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{C})~\cite{mobley2007predicting,aldeghi2015accurate,aldeghi2017predictions}; 
\item small alchemical modifications can be made between chemically related ligands to estimate relative differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{E})~\cite{wang2015accurate,mey2016blinded,song2019using,gapsys2020large,kuhn2020assessment}.
\end{itemize}

After an alchemical calculation is performed, which frequently involves multiple simulations at a variety of alchemical states, the data must be analyzed to compute an estimate of the free energy for the transformation of interest.
Early work used simple but statistically suboptimal estimators for this: free energy perturbation (FEP) uses a simple (but highly biased) estimator based on the Zwanzig relation~\cite{zwanzig1954hightemperature} or numerical quadrature via thermodynamic integration (TI), for which the theory dates back the better part of a century but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamics, merz1989free}. Macuglia et al. recently reviewed the evolution of biomolecular alchemical free energy calculations into practically useful tools \cite{macuglia2025dawn}. %
More recent developments have seen new, highly efficient statistical estimators that make better use of all the data, often building on the more efficient and less biased Bennett acceptance ratio (BAR)~\cite{bennett1976efficient}, producing multistate generalizations~\cite{shirts2008statisticallya} or removing the need for global equilibrium~\cite{wu2016multiensemble, mey2014xtram, wu2014statistically}.

@@

As the field of molecular simulation can now routinely access microsecond timescales with the aid of GPUs~\cite{salomon-ferrer2013routine}, and millisecond timescales appear to soon be within reach, accurate alchemical calculations on even more challenging problems will become reasonable to perform. 
In part because of the many advances that have been made in developing these important computational approaches, today's users may find it difficult to get started with these complex calculations while also keeping up with the fast pace of change. 
This ``Best Practices'' guide aims to provide current recommendations and tips for users of all experience levels, providing context with other new methods being developed. Updates and suggestions are highly encouraged via our GitHub repository at \url{https://github.com/alchemistry/alchemical-best-practices}.

@@

This ``Best Practices'' guide focuses on providing a good starting point for new practitioners and a reference for experienced practitioners. 
For this purpose we provide a convenient checklist (Sec.~\ref{sec:checklist}) to help ensure all calculations can take advantage of currently-understood best practices for alchemical simulation and analysis. Where the best practices are currently not certain, we highlight areas where further research is needed to identify an unambiguous recommendation.
This guide can also serve as a set of best practices to ensure simulation robustness and reproducibility which reviewers may wish to consider as they evaluate papers including alchemical free energy calculations.

In this document, we assume that novice practitioners have at least moderate experience with molecular simulation concepts and use of simulation packages. 
We also assume a basic familiarity with the principles of molecular mechanics, molecular dynamics simulations, statistical mechanics, and the biophysics of protein-ligand association. If you feel unfamiliar with some of these concepts, good starting points can be found in these references~\cite{braun2019best, grossfield2018best, klimovich2015guidelines, shirts2012best}.

 While reading this ``Best Practices'' guide, it is important to bear in mind \emph{this is not a review} of all free energy calculation methods at the cutting edge of current research.
Instead this guide aims to answer the following questions:
\begin{itemize}
    \item Is my problem suitable for an alchemical calculation? 
     \item How do I prepare my system to carry out an alchemical calculation?
     \item How do I select an appropriate simulation protocol? 
     \item What software tools are available to perform alchemical calculations? 
     \item How should I analyze my data and report uncertainties? 
\end{itemize}

Notably, this document---version 2.0 of the original guide---focuses specifically on best practices on single alchemical free energy calculations. A companion guide, \textit{Best Practices for Multiple Alchemical Free Energy Calculations}, covers the considerations involved in high-throughput or large-scale campaigns, such as automation, perturbation network design, and campaign-level data management and analysis. Readers aiming to scale up to multiple calculations will find many important suggestions and ideas relevant to those problems in that document.

In this guide, some additional background information may be needed depending on the nature of the alchemical project. For example, if binding poses are not known, docking calculations or co-folding models can be used to generate an initial small molecule binding pose to start alchemical simulations\cite{scheen2025leveraging, thaler2025boltz}. This will require some basic familiarity with how to perform docking or use generative models to generate reasonable simulation starting points~\cite{grinter2014challenges}. 
As some of the theoretical background can be daunting to beginners, we do, however, provide a guide to the essential theory behind alchemical free energy calculations in Sec.~\ref{sec:theory}.

In the remainder of this paper, we will cover topics that are key to the preparation (Sec.~\ref{sec:prerequisites}), choice and use of correct protocols (Sec.~\ref{sec:simulation_protocol_choice}), and finally the best practices that should be used in the analysis of alchemical calculations (Sec.~\ref{sec:data_analysis}). 
We will give particular focus to aspects of the molecular simulations which are unique to alchemical calculations---these include the calculation of transfer free energies (hydration free energies, partition coefficients, etc.), and binding free energies (absolute and relative). We primarily focus on free energy calculations using simulations performed at \emph{equilibrium} in this Best Practices guide, as best practices for these are more developed, and non-equilibrium techniques may warrant their own guide as such practices evolve.

While we try to address as many methods and practices as possible, the field of free energy calculations is broad, and there are many advanced topics that are left to future ``Best Practices documents'' focusing on specific issues. 
Below, we provide a non-exhaustive list of topics we have \emph{not} addressed, along with some references to provide starting points on these more advanced topics:
\begin{itemize}
\item covalent inhibition~\cite{lameira2019predicting}
\item free energies of mutation of protein side chains~\cite{gapsys2016accurate,aldeghi2018accurate}
\item nonspecific binding or multiple binding sites~\cite{gill2018binding}
\item approximate and often less accurate endpoint free energy methods such as MM-PBSA~\cite{genheden2015mm} and LIE~\cite{gutierrez-de-teran2012linear}
\item Free energy methods that extract the ligand using geometric order parameters and potential of mean force methods~\cite{heinzelmann2017attachpullrelease}
\item Forcefield dependence for protein, ligand, ions, co-solvents, and co-factors. A number of different studies have looked at the influence of force fields and it is assumed the user has made an appropriate choice for the system under study~\cite{loeffler2018reproducibility, vassetti2019assessment, lopes2015current}. 
\item Non-equilibrium free energy calculations~\cite{gapsys2020large}
\item Free energy calculations using QM/MM methods~\cite{beierlein2011simple,dybeck2016comparison,cave-ayland2015direct,Schoeller2023}.
\item Free energy calculations using machine learning methods~\cite{rufa2020chemical, scheen2020hybrid, cole2020machine,Karwounopoulos2024}
\end{itemize}

@@

\begin{tcolorbox}[title=List of Symbols, colback=green!10!white]
$A$ --- Helmholtz free energy (free energy in the canonical ensemble) or Helmholtz function, with Helmholtz free energy used in the text.\\
$b$ --- reduced binding energy function \\
$C^{\circ}$ --- standard state concentration \\
$C_t$ --- discrete-time-normalized fluctuation auto-correlation function\\
$f$ --- reduced (dimensionless) free energy \\
$G$ --- Gibbs free energy (free energy in the isothermal isobaric ensemble), Gibbs function, or free enthalpy, though the most common term Gibbs free energy is used in the text\\
$g$ --- statistical inefficiency\\
$h$ --- Planck's constant \\
$\vec{h}$ --- a displacement vector in the ATM method
$k_B$ --- Boltzmann constant \\
$k_H$ --- Henry's constant \\
$K_b^{\circ}$ --- binding  constant \\
$L$ and $R$ --- generic labels for ligand and receptor\\
$\mathcal{O}$ --- overlap matrix between alchemical states\\
$P^{\circ}$ --- standard pressure \\
$p$ --- pressure \\
$\vec{q}$ --- vector of a single configuration, i.e. $x$, $y$, $z$ coordinates of the simulation system\\
$T$ --- temperature \\
$t_0$ --- equilibration time \\
$U$ --- potential energy\\
$u$ --- reduced (dimensionless) potential describing a thermodynamic state \\
$w(b, \vec{\lambda})$ --- perturbation energy function \\
$Z$ --- partition function \\
$\beta$ --- inverse thermal energy $(k_B T)^{-1}$ \\
$\Gamma$ --- configurational space accessible by simulations \\
$\Delta \hat{f}$ --- estimate from an estimator for the reduced free energy difference between two states\\
$\vec{\lambda}$ --- alchemical progress parameter, which may be multidimensional \\
$\mu$ --- chemical potential\\
$\tau _{eq}$ --- integrated auto-correlation time\\
\end{tcolorbox}

@@

In this section, we use the statistical mechanics theory of dilute solutions to derive free energy expressions that form the basis of alchemical computational protocols to model non-covalent molecular binding, solvation, and solvent partitioning equilibria. The emphasis here is placed on bridging theoretical foundations and intuition. Since chemical equilibria are regulated by the chemical potentials of the species involved, we will start by considering a statistical mechanics expression of the chemical potential of a solute in a solvent. We will then combine chemical potentials to develop models for each process.

@@

The chemical potential of a solute $u$ in an ideally diluted solution in a solvent $v$ can be written as~\cite{guggenheim1952mixtures504,gilson1997statisticalthermodynamic,gilson2007calculation}
\begin{equation}
  \mu_u = -k_B T \ln \frac{q'_u}{C_u \Lambda_u^3}
  \label{eq:chemical-potential-definition}
\end{equation}
where $k_B$ is Boltzmann's constant, $T$ is the absolute temperature, $C_u = N_u/V$ is the number concentration of the solute, $\Lambda_u = h/\sqrt{2 \pi M_u k_B T}$ is the thermal de Broglie wavelength of the solute, where $h$ is Planck's constant and $M_u$ is the solute's total mass. The quantity $q'_u$ in Eq.~(\ref{eq:chemical-potential-definition}) is the molecular partition function of the solute at rest in the solvent
\begin{equation}
  q'_u = \frac{\Lambda_u^3}{\prod_i \lambda_i^3} 8 \pi^2 z_u
  \label{eq:intra-qpu-def}
\end{equation}
where $\beta = k_B T)^{-1}$, $\lambda_i$ is the thermal De Broglie wavelength of the $i$-th atom of the solute (not the alchemical parameter as used elsewhere in this document), and $z_u$ is the intramolecular configurational partition function of the solute
\begin{equation}
  z_u = \int dx_u e^{-\beta \Psi_v(x_u)}
  \label{eq:intra-zu-def}
\end{equation}
where $x_u$ represents the collection of internal degrees of freedom of the solute (bond lengths, bond angles, and dihedral angles) with the corresponding volume element $dx_u$ assumed to include the appropriate Jacobian factors. The function $\Psi_v(x_u)$ in Eq.~\ref{eq:intra-zu-def} is the solvent-averaged potential of mean force of the solute in conformation $x_u$ in solvent $v$\cite{rouxandsimonson1999solventmodels} 
\begin{equation}
  e^{-\beta \Psi_v(x_u)} = \frac{1}{Z_{N_v}} \int d\vec{r}_v e^{-\beta U(x_u, \vec{r}_v)}
  \label{eq:solvent-pmf-def}
\end{equation}
where $\vec{r}_v$ represents the collection of Cartesian coordinates of the atoms of $N_v$ solvent molecules, $U(x_u, \vec{r}_v)$ is the potential energy of the solvent with one solute molecule in conformation $x_u$ placed in an arbitrary position in a volume $V$. When the potential energy function of the solution is separable into an intramolecular potential energy of the solute $U(x_u)$ and a solute-solvent non-bonded interaction energy, the potential of mean force can be written as the sum $\Psi_v(x_u) = U(x_u) + W_v(x_u)$ where $W_u(x_u)$ is the solvent potential of mean force, the solvation free energy of the solute fixed in configuration $x_u$\cite{gilson2007calculation}. However, the potential of mean force definition Eq.~(\ref{eq:solvent-pmf-def}) is valid in general. Finally, $Z_{N_v}$ is the configurational partition function of the pure solvent
\begin{equation}
  Z_{N_v} = \int d\vec{r}_v e^{-\beta U(\vec{r}_v)} \, .
  \label{eq:Z-pure-solvent}
\end{equation}

A few notes of explanation are in order. First, Eq.~(\ref{eq:chemical-potential-definition}) is derived using classical statistical mechanics theory~\cite{hill1986statthermo}. Classical mechanics is considered an excellent approximation for the room temperature processes studied here involving only changes in intermolecular interactions. A quantum mechanical treatment is necessary for processes not covered here, such as covalent binding, that involve breaking and forming chemical bonds. Classically, equilibrium thermodynamics properties, such as gas solubilities and binding constants, do not depend on atomic masses. Hence, while they are included in Eqs.~\ref{eq:chemical-potential-definition} and Eq.~\ref{eq:intra-qpu-def} for completeness, the thermal De Broglie wavelength terms coming from the partition functions of the momenta cancel in the resulting expressions below and we will no longer consider them.

Eq.~(\ref{eq:chemical-potential-definition}) is derived by writing the classical canonical configurational partition function $Z(N_u = 1, N_v)$ of one solute molecule in a solvent of $N_v$ molecules in a volume $V$ and multiplying and dividing it by configurational partition function of the pure solvent [Eq.~(\ref{eq:Z-pure-solvent})], and then separating out the integration over the external degrees of freedom of the solute (translations and rotations) giving an $8 \pi^2 V$ term to yield $Z(N_u = 1, N_v) = Z_{N_v} 8 \pi^2 V z_u$ where $z_u$ is defined by Eqs.~(\ref{eq:intra-zu-def}) and (\ref{eq:solvent-pmf-def}). The canonical configurational partition function of $N_u$ independent solute molecules in a solvent is then obtained by considering the product of the solute terms $Z(N_u, N_v) = Z_{N_v} (8 \pi^2 V z_u)^{N_u}/N_u!$,~\cite{guggenheim1952mixtures504,simonson2016physical} where the factorial term accounts for indistinguishability. Finally, the canonical partition function $Q(N_u, N_v)$ of the solution is obtained by including the appropriate terms coming from the momenta and differentiating the resulting Helmholtz free energy $A(N_u, N_v) = -k_B T \ln Q(N_u, Nv)$ with respect to $N_u$ to obtain Eq.~(\ref{eq:chemical-potential-definition}).

@@

We derived Eq.~(\ref{eq:chemical-potential-definition}) using the canonical (constant volume) partition function of the solution. Since the chemical potential can be equivalently obtained from the constant-volume differentiation of the Helmholtz free energy or the constant-pressure differentiation of the Gibbs free energy, Eq.~(\ref{eq:chemical-potential-definition}) is correct. However, the volume dependence of the chemical potential has to be considered when modeling the Gibbs free energy change for constant-pressure processes where the system's volume changes significantly~\cite{gilson1997statisticalthermodynamic}. Terms due to volume variations at constant pressure are generally considered small for most processes of solvation, solvent partitioning, and molecular binding at atmospheric pressure and are often ignored in alchemical calculations.  This is because when a single molecule is eliminated, errors will be of the magnitude of the fractional change of the system size.  When one particle in 1000 disappears, this will be negligible with respect to other source of error. Hence, we will use Eq.~(\ref{eq:chemical-potential-definition}) to derive expressions for $\Delta G$'s.
